---
title: "SteppedPower"
author: "Philipp Mildenberger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stepped Power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(knitr)
library(SteppedPower)
library(Matrix)
```

## About `SteppedPower`

The package `SteppedPower` aims at providing tools for power calculation 
and design diagnostics in interventional study designs,
with a special focus on stepped wedge designs.  

This code is modularised in order to be flexible and easy to use 
(and hopefully to maintain as well). 
At the same time, the use of sparse matrix classes of the wonderful `matrix` 
package makes computation of large designs feasible.


## A quick tour

For most users, the probably most important function is `wlsMixedPower`. 
It calls several constructor functions that will be shortly discussed in the next 
section.

To define the design experimental design, 

```{r}
#wlsMixedPower()
```

## The 

It calls `compute_wlsPower`, which is the workhorse function of this package.

### Define Study and Analysis Design

`construct_DesMat` 

`construct_trtMat`

`construct_timeadjust`
If not specified, the number of timepoints is guessed as the fewest number of periods (timepoints) possible with the given design;

### Define assumed Covariance Structure

`construct_CovMat`


### Gimmicks

`plot` method



## Use cases

### Comparison of two groups -- Z-Test

The variance is assumed to be known, 
Lets first start with a rather simple example where we want to compare the 
mean in two groups. Lets say we have two groups of 10 observations

The Z-test can be expressed as follows:

```{r}
wlsMixedPower(Cl=c(10,10), sigma=1, mu0=0,mu1=.6, N=1, design="parallel", timepoints=1)

## the same:
wlsMixedPower(Cl=c(1,1), sigma=1, mu0=0,mu1=.6, N=10, design="parallel", timepoints=1)
```


```{r}
pwr::pwr.norm.test(.3,n=20)
```

>A quick Note on t-tests: It is much more challenging to us `SteppedPower` to reproduce settings in 
which the variance is assumed to be unknown, most prominently the well known t-test.
In this package, you find implemented some heuristics for guessing the denominator 
degrees of freedom, but they yield rather scaled Wald test than t tests.
The main difference is that the distribution under the alternative is assumed to be symmetric,
whereas the t-test assumes a non-central (hence skewed) t-distribution.

### Comparison of two groups -- t-Test

Calculating power 
```{r}
a <- wlsMixedPower(Cl=c(10,10), sigma=1, mu0=0,mu1=.6, N=1, dfAdjust="between-within",
                   timepoints=1, design="parallel", verbose = TRUE)
# a$CovarianceMatrix
dsnmatrix <- a$DesignMatrix$dsnmatrix
  tmpmat <- t(dsnmatrix) %*% Matrix::chol2inv(Matrix::chol(a$CovarianceMatrix))
  VarMat <- Matrix::solve(tmpmat %*% dsnmatrix)
VarMat[1,1]

pwr::pwr.t.test(10,.6, type="two.sample")
pwr::pwr.t.test(20,.3, type="one.sample")
```

### Longitudinal study -- parallel groups


```{r}
wlsMixedPower(Cl=c(10,10),timepoints=5,mu0=0,mu1=.25,sigma=.5,design="parallel")
wlsMixedPower(Cl=c(10,10),timepoints=5,mu0=0,mu1=.25,sigma=.5,tau=.2,design="parallel")
```

### Stepped Wedge designs with empty sequences (i.e. waves)

Periods in which no cluster switches to the intervention are specified by inserting zeros into 
the `Cl` argument, i.e. `Cl=c(4,4,4,0)`.

```{r}
mod3 <- wlsMixedPower(Cl=c(4,4,4,0), sigma=2, eta=1, mu0=0,mu1=1)
mod4 <- wlsMixedPower(Cl=c(4,4,4),   sigma=2, eta=1, mu0=0,mu1=1, timepoints=5)

mod1 <- wlsMixedPower(Cl=c(0,4,4,4), sigma=2, eta=1, mu0=0,mu1=1)
mod2 <- wlsMixedPower(Cl=c(4,4,4,0), sigma=2, eta=1, mu0=0,mu1=1, trtDelay=0, verbose=TRUE)
```


### Incomplete Stepped Wedge Designs

Suppose you do not plan to observe all clusters over the whole study period.
Rather, clusters that switch early to the intervention are not obeserved until the end.
Analogous, observation starts later in clusters that switch towards the end of the study.
This is sometimes called 'incomplete SWD'.  

There are two ways to achieve this in `SteppedPower`, both by using the `incomplete` argument.
One can either scalar, which then defines the number of observed periods before and 
after the switch from control to intervention in each cluster.

If for example the study consists of eight clusters in eight waves (i.e. nine timepoints), 
and we observe four timepoints before and after the switch, then we receive

```{r}
wlsPwr <- wlsMixedPower(Cl=rep(1,8), sigma=1, tau=.6, mu0=0,mu1=1, incomplete=4, verbose=TRUE)
wlsPwr
```

A slightly more tedious, but more flexible way is to define  a matrix where the number of rows corresponds 
to the total number of clusters and the number of columns corresponds to the number of timepoints. 
The value in each cell is the residual standard deviation, if a cluster is not observed at a specific timepoint, 
set the value in the corresponding cell to `0`. For the example above, such a matrix would look like this:
```{r}
TM  <- toeplitz(c(1,1,1,1,0,0,0,0))
incompleteMat <- cbind(TM[,1:4],rep(1,8),TM[,5:8])
incompleteMat
```

Now all that's left to do is to plug that into the main function:
```{r}
wlsPwr <- wlsMixedPower(Cl=rep(1,8), sigma=1, tau=.6, EffSize=1, incomplete=incompleteMat, verbose=TRUE)
wlsPwr
```

We can also have a quick look at the projection matrix where we see that the 
clusters have a weight of exactly zero at the timepoints where they are not observed
```{r}
plot(wlsPwr)
```


### non-standard time effects

The by far most usual method for the modeling of potential secular trends is
to take time period as a factor into the analysis model. This is also proposed by Hussey&Hughes2007, Hemming2015? ...

In some scenarios, different modeling of the time trend can be favoured. 

```{r}
timeMat <- matrix(c(1,1,1,1,1,1,1,1,
                    0,1,0,0,0,1,0,0,
                    0,0,1,0,0,0,1,0,
                    0,0,0,1,0,0,0,1),8,4)
DesMat  <- construct_DesMat(rep(1,7), timeBlk = timeMat)

wlsPower <- wlsMixedPower(DesMat=DesMat, mu0=0,mu1=.8, sigma=1, tau=.2, verbose=TRUE)
plot(wlsPower)
wlsPower <- wlsMixedPower(Cl=rep(1,7), mu0=0,mu1=.8, sigma=1, tau=.2, verbose=TRUE)
plot(wlsPower)
```



### Closed cohort SWD

This requires quite a bit of configuration that will be incorporated into the package in the future.

Here we have just a simple example as a sketch. 20 clusters in 4 waves. 
50 Individuals within each cluster.
```{r}
Cl <- c(5,5,5,5)
nInd <- 50
tp <- length(Cl)+1
psi <- .1 ; sigma <- .3 ; tau=.5

Ind    <- construct_CovBlk(timepoints=tp, sigma=.3, tau=.5)
CovBlk <- bdiag(replicate(nInd,Ind,simplify=FALSE)) + matrix(psi^2,nrow=5*nInd,ncol=5*nInd)
CovMat <- construct_CovMat(CovBlk=CovBlk,SumCl=sum(Cl))

wlsMixedPower(Cl=c(5,5,5,5)*50, CovMat=CovMat, mu0=0,mu1=0.05, verbose=TRUE)
wlsMixedPower(Cl=c(250,250,250,250),sigma=.3,tau=.5,mu0=0,mu1=.05)

sigma<-.3 ; tau <- .5 ;n <- 50
sqrt(sigma^2+1/n*tau^2)

wlsMixedPower(Cl=Cl, sigma=.3/sqrt(50), tau= .5/sqrt(50), mu0=0,mu1=.05)
wlsMixedPower(Cl=Cl, sigma=.3, tau= .5/sqrt(50), N=50, mu0=0,mu1=.05)

wlsMixedPower(Cl=Cl, sigma=.3/sqrt(50), tau= sqrt((.5/sqrt(50))^2), mu0=0,mu1=.05, verbose = TRUE)
tmp <- wlsMixedPower(Cl=Cl, sigma=.3/sqrt(50), tau= sqrt((.5/sqrt(50))^2+(psi)^2), mu0=0,mu1=.05, verbose = TRUE)
tmp$CovarianceMatrix[1:5,1:5]

CM50 <- CovMat[1:50,1:50]
sum(CM50^2)/2500

(diag(CM50))/(50)
```

> runtime is O(timepoints^2 nInd^2) , but O(sum(Cl)) 


```{r}

## tau is random effect in subcluster, psi is random effect in Cluster
## 

Cl <- c(5,5,5,5)
nInd <- 50
tp <- length(Cl)+1
psi <- .2 ; sigma <- .3 ; tau=.3 ; eta <- .4 ; rho <- 0
DesMat <- construct_DesMat(Cl=Cl)
etaMat <- eta * DesMat$trtMat
etaLst <- split(etaMat, row(etaMat)) 

# debugonce(construct_CovBlk)
Ind_to_CovBlk <- function(etas,Cl_i){
  Ind    <- construct_CovBlk(timepoints=tp, sigma=sigma, tau=tau, eta=etas)
  CovBlk <- bdiag(replicate(nInd,Ind,simplify=FALSE)) + matrix(psi^2,Cl_i*nInd,Cl_i*nInd)
}

CovMat <- bdiag(mapply(Ind_to_CovBlk, etas=etaLst, Cl_i=Cl))
wlsMixedPower(Cl=c(5,5,5,5)*50, CovMat=CovMat, mu0=0,mu1=0.05)

wlsMixedPower(Cl=Cl, sigma=.3/sqrt(50), tau= sqrt((tau/sqrt(50))^2+(psi)^2), eta=eta/sqrt(50), mu0=0,mu1=.05)

## with rho
b3 <- wlsMixedPower(Cl=Cl, sigma=.3/sqrt(50), tau= sqrt((tau/sqrt(50))^2+(psi)^2), eta=eta/sqrt(50), 
                    rho=rho, mu0=0,mu1=.05, verbose=TRUE)


# without random cluster effect (only random subcluster effect)
b1 <- wlsMixedPower(Cl=Cl, sigma=.3/sqrt(50), tau= (.5/sqrt(50)), eta=.4/sqrt(50), 
                    mu0=0,mu1=.05, verbose = TRUE)

plot(b3)
b3$CovarianceMatrix[1:5,1:5]

```



















