---
title: "SteppedPower"
author: "Philipp Mildenberger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stepped Power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## About `SteppedPower`

The package `SteppedPower` aims at providing tools for power calculation 
and design diagnostics in interventional study designs,
with special focus on stepped wedge designs.  

This code is highly modularised in order to be flexible and easy to use 
(and hopefully maintain as well). 
At the same time, the use of sparse matrix classes of the wonderful `matrix` 
package makes computaion fast (and numerically stable).


## A quick tour

For most users, the probably most important function is `wlsMixedPower`. 
It calls `compute_wlsPower`, which is the workhorse function of this package.

### Define Study and Analysis Design

`construct_DesMat` 


### Define assumed Covariance Structure

`construct_CovMat`


### Gimmicks

`plot` method






## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

