---
title: "Getting Started with `SteppedPower`"
author: "Philipp Mildenberger"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: true
    theme: lumen
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{GettingStartedWithSteppedPower}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
library(knitr)
library(SteppedPower)
library(Matrix)
```

# About `SteppedPower`

The package `SteppedPower` aims at providing tools for power calculation 
and design diagnostics in interventional study designs,
with a special focus on stepped wedge designs. Other implemented study design types are 
parallel, parallel with baseline period(s) and crossover designs. Further design types can be 
easily defined by the user.  

Currently, normal outcomes and binomial outcomes with logit link are implemented.  
Random cluster effect, autocorrelated cluster effect, random treatment effect, 


This code is modularised in order to be flexible and easy to use 
(and hopefully to maintain as well). 
At the same time, the use of sparse matrix classes of the wonderful `matrix` 
package makes computation of large designs feasible. 
<!-- [@hussey2007design] -->

# Methods and Notation

A common approach to model the correlation in longitudinal studies are random effects.
Such a model has the form


$$y_{ijk}= \mu_0 + \alpha_i + T_{ij} \theta + b_j + e_{ijk}$$ 

with

* $y_{ijk}$ the response in cluster $j$ at time $i$ for individual $k$
* $\mu_0$ a overall mean (under control)
* $\alpha_i$ the time trend at time $i$
* $T_{ij}$ indicates the treatment status (0 = control, 1 = interventional treatment)
* $\theta$ the treatment effect 
* $b_j$ a random cluster effect for cluster $j$ with $b_j \sim N(0,\tau^2)$
* $e_{ijk}$ a normal random error term with $e_{ijk}\sim N(0,\sigma^2)$


For power calculation, the standard deviation of random effects is assumed to be known.
Lets define $\beta:=(\mu,\alpha',\theta)'$ and $\omega_{ijk}:=b_j+e_{ijk}$. 
This leads to a compact and more general notation of the above equation:

$$\begin{align}
y_{ijk}&= X_{ij}\beta + \omega_{ijk}\\
\text{or, in matrix notation:} \qquad \\
y&=X\beta + \omega
\end{align}$$ 

Where $X$ is the corresponding design matrix and $\omega\sim N(0,\Omega)$,
where $\Omega$ is a compound-symmetry (syn. exchangeable) variance matrix defined 
by $\tau$ and $\sigma$. We are thus in a weighted least squares setting, 
so the variance of $\beta$ is

$$ \text{Var}(\hat\beta) =  {(X'\Omega^{-1}X)^{-1}}$$ 


We can then calculate the power of a z-test 

$$ \text{power} = \Phi\left(\frac{\theta_A-\theta_0}{\sqrt{\text{Var}(\hat \theta)}}- Z_{1-\frac{\alpha}{2}}\right) $$ 
<!-- \eqn{\text{power} = \Phi\left(\frac{\theta_A-\theta_0}{\sqrt{\text{Var}(\hat \theta)}}- Z_{1-\frac{\alpha}{2}}\right) } -->


where $\text{Var}(\hat \theta)$ is the diagonal element of $\Omega$ 
that corresponds to $\hat\theta$.
 

  
  
Extensions to the above formula implemented in this package are  

  * random treatment effect
  * autoregressive cluster effect
  * binomial outcomes with logit-link
  

$$y_{ijk}= g\big( \mu + \alpha_i + X (\theta_{ij} + c_j) + b_j + e_{ijk}\big)$$ 
with  

  * $g(\cdot)$ a link function
  * $c_j$ a random treatment effect
  * $b_j$ and $c_j$ jointly distributed with 
  $\left(\begin{smallmatrix} \tau^2 & \rho\tau\eta \\ \rho\tau\eta & \eta^2\end{smallmatrix}\right)$


# A quick tour

For most users, the probably most important function is `wlsPower`. 
It calls several auxiliary functions which will be shortly discussed here.
This section is not essential for the usage of `SteppedPower`, it might be
helpful to design non-standard userdefined settings. 

`wlsPower` is essentially just a flexible wrapper for the function 
`compute_wlsPower`, which does the actual computation.  

`construct_DesMat` builds the design matrix which consists of the treatment 
status, usually built by`construct_trtMat` and the time adjustment, usually
built by `construct_timeadjust`. 
There is also the option to pass a user defined definition of the treatment 
status to `construct_DesMat`.
If not specified, the number of timepoints is guessed as 
the fewest number of periods (timepoints) possible with the given design, 
i.e. two for cross-over designs or the number of waves plus one
for stepped wedge designs. 


`construct_CovMat` builds the covariance matrix (explicitly). It uses 
`construct_CovBlk` to construct the blocks for each cluster which are then combined
to a block diagonal matrix.

```{r}
CovBlk <- construct_CovBlk(sigma=c(1,1), tau=c(.5,.5))
all.equal(CovBlk,diag(2) + c(.5,.5) %o% c(.5,.5))

CovMat1 <- construct_CovMat(CovBlk = CovBlk, SumCl=2)
CovMat2 <- construct_CovMat(SumCl=2,timepoints=2,sigma=1,tau=.5)

CovMat1
all.equal(CovMat1,CovMat2)
```



# Features 

## Plot Method
<!-- TODO: NAMING THINGS!! how to call the influence matrix??? -->
In the weighted least squares setting, the estimator $\hat \beta$ is a linear function of the data $y$
$$\hat\beta = \underbrace{(X'\Omega^{-1}X)^{-1}(X'\Omega^{-1})}_{=:\text{M}}\cdot y$$
with $X$ the design matrix and $\Omega$ the covariance matrix as above.
The matrix $M$ gives an impression of the importance of clusters and 
time periods with regard to the estimated coefficients $\hat\beta$. 
The first row of $M$ corresponds to the coefficient of the treatment status, 
i.e. the treatment effect.  
The `plot.wlsPower` method visualises this first row of $M$ as a matrix 
where rows and columns correspond to clusters and time periods, respectively.

Furthermore, to give a rough comparison of importance between clusters 
(or between time periods), the sum of absolute weights per row (or per column) 
is also shown. 

>CAVE: These are out-of-bag estimates for the influence of observations on 
$\hat\theta$, but not for $\text{Var}(\hat\theta)$ !

```{r}
wlsPwr <- wlsPower(Cl=c(3,2,3), mu0=0, mu1=1, sigma=1, tau=.5, verbose=2)
plot(wlsPwr)
```

## Find Sample Size for given Power

When the argument `Power` is passed to `wlsPower`, 
the sample size needed is calculated, under the assumption of equally sized 
clusters and periods. 
```{r}
wlsPower(Cl=c(3,2,3), mu0=0, mu1=.2, sigma=1, tau=0, Power=.8)
```



# Use cases and examples

## Comparison of two groups -- Z-Test

This might be a proof of concept rather than an example with practical relevance. 
We want to compare the mean in two groups. Lets say we have two groups of 10 observations

The Z-test can be expressed as follows:

```{r}
wlsPower(Cl=c(10,10), mu0=0,mu1=.6,sigma=1, tau=0, N=1, 
              dsntype="parallel", timepoints=1)

## the same:
wlsPower(Cl=c(1,1), mu0=0,mu1=.6, sigma=1, tau=0, N=10,
              dsntype="parallel", timepoints=1)
```


```{r}
pwr::pwr.norm.test(.3,n=20)$power
```

>A quick Note on t-tests: It is much more challenging to use `SteppedPower` to reproduce settings in 
which the variance is assumed to be unknown, most prominently the well known t-test.
In this package, you find implemented some (experimental) heuristics 
for guessing the denominator degrees of freedom, but they yield rather 
scaled Wald tests than t tests. 
The main difference is that the distribution under 
the alternative is assumed to be symmetric,
whereas the t-test assumes a non-central (hence skewed) t-distribution.


## Longitudinal study -- parallel groups


```{r}
wlsPower(Cl=c(10,10),timepoints=5,mu0=0,mu1=.25,sigma=.5,dsntype="parallel")
wlsPower(Cl=c(10,10),timepoints=5,mu0=0,mu1=.25,sigma=.5,tau=.2,dsntype="parallel")
```

<!-- ### Comparison of designs -->

<!-- See Hemming, Haines 2014 -->

<!-- $$ icc = \frac{tau^2}{sigma^2+tau^2}$$  -->

<!-- $$  \frac{1}{icc} = \frac{sigma^2+tau^2}{tau^2} = \frac{sigma^2}{tau^2} + 1  $$  -->

<!-- $$  \frac{1-icc}{icc} = \frac{sigma^2}{tau^2}  $$  -->

<!-- $$ \tau^2 = \sigma^2 \frac{icc}{1-icc}$$  -->

<!-- ```{r, eval=FALSE} -->
<!-- library(swCRTdesign) -->

<!-- # icc <- 0.1 -->
<!-- # tau001 <- sqrt(icc/(1-icc)) -->
<!-- # tau01 <- sqrt(icc/(1-icc)) -->

<!-- swPwr(swDsn(c(5,5,5,5)), distn="gaussian", n=50, mu0=0, mu1=.3, sigma=1, tau=0, eta=0, rho=0, gamma=0, retDATA = TRUE) -->
<!-- wlsPower(Cl=c(5,5,5,5), mu0=0, mu1=0.3, sigma=1, N=50, tau=0) -->

<!-- swPwr(swDsn(c(5,5,5,5)), distn="gaussian", n=50, mu0=0, mu1=.3, sigma=1, tau=tau001,  -->
<!--       eta=0, rho=0, gamma=0, retDATA = TRUE) -->
<!-- wlsPower(Cl=c(5,5,5,5), mu0=0, mu1=0.3, sigma=1, N=50, tau=tau001) -->
<!-- swPwr(swDsn(c(5,5,5,5)), distn="gaussian", n=50, mu0=0, mu1=.3, sigma=1, icc=0.01, cac=1) -->

<!-- swPwr(swDsn(c(5,5,5,5)), distn="gaussian", n=50, mu0=0, mu1=.3, sigma=1, tau=tau01,  -->
<!--       eta=0, rho=0, gamma=0, retDATA = TRUE) -->
<!-- wlsPower(Cl=c(5,5,5,5), mu0=0, mu1=0.3, sigma=1, N=50, tau=tau01) -->
<!-- swPwr(swDsn(c(5,5,5,5)), distn="gaussian", n=50, mu0=0, mu1=.3, sigma=1, icc=0.1, cac=1) -->

<!-- swPwr(swDsn(c(5,5,5,5)), distn="gaussian", n=50, mu0=0, mu1=.3, sigma=1, tau=100,  -->
<!--       eta=0, rho=0, gamma=0, retDATA = TRUE) -->
<!-- wlsPower(Cl=c(5,5,5,5), mu0=0, mu1=0.3, sigma=1, N=50, tau=100) -->
<!-- swPwr(swDsn(c(5,5,5,5)), distn="gaussian", n=50, mu0=0, mu1=.3, sigma=1, icc=.1, cac=0) -->


<!-- wlsPower(Cl=c(10,10), mu0=0, mu1=0.3, sigma=sqrt(1/9+.02), dsntype="parallel") -->
<!-- wlsPower(Cl=c(10,10), mu0=0, mu1=0.3, sigma=sqrt(1/9+.02), dsntype="parallel_baseline") -->

<!-- wlsPower(Cl=c(10,10), mu0=0, mu1=0.3, sigma=1, N=50, tau=tau001, dsntype="parallel") -->
<!-- wlsPower(Cl=c(10,10), mu0=0, mu1=0.3, sigma=1, N=50, tau=tau001, dsntype="parallel_baseline") -->

<!-- wlsPower(Cl=c(10,10), mu0=0, mu1=0.3, sigma=1, N=50, tau=tau01, dsntype="parallel") -->
<!-- wlsPower(Cl=c(10,10), mu0=0, mu1=0.3, sigma=1, N=50, tau=tau01, dsntype="parallel_baseline") -->

<!-- swPwr(swDsn(c(5,5,5,5)), distn="gaussian", n=50, mu0=0, mu1=.3, sigma=1, icc=0.1, cac=1, retDATA=TRUE) -->
<!-- wlsPower(Cl=c(5,5,5,5), mu0=0, mu1=0.3, sigma=sqrt(1/9+.02), N=1, tau=0) -->

<!-- View(swPwr) -->
<!-- ``` -->



## Stepped Wedge designs with empty sequences (i.e. waves)

Periods in which no cluster switches to the intervention are specified by inserting zeros into 
the `Cl` argument, i.e. `Cl=c(4,4,4,0)`.

```{r}
mod3 <- wlsPower(Cl=c(4,4,4,0), sigma=2, eta=1, mu0=0,mu1=1)
mod4 <- wlsPower(Cl=c(4,4,4),   sigma=2, eta=1, mu0=0,mu1=1, timepoints=5)

mod1 <- wlsPower(Cl=c(0,4,4,4), sigma=2, eta=1, mu0=0,mu1=1)
mod2 <- wlsPower(Cl=c(4,4,4,0), sigma=2, eta=1, mu0=0,mu1=1, trtDelay=0)
```

## Autocorrelated cluster effects

In longitudinal studies, it can be sensible to assume that correlation within 
clusters decreases with increasing time lag.
The argument `tauAR` enables the user to specify a AR-1 correlation. `tauAR`
must be any value between `0` and `1`. The former corresponds to i.i.d. observations, 
the latter to the usual compound symmetry covariance type.  

An example of a stepped wedge design with 8 clusters in 4 waves, once with 
medium autocorrelation (`tauAR=0.6`) and once with high autocorrelation (`tauAR=0.9`):
```{r}
a <- wlsPower(Cl=c(2,2,2,2), mu0=0, mu1=1, sigma=1, N=100, tau=1, tauAR=.6, verbose=2)
b <- wlsPower(Cl=c(2,2,2,2), mu0=0, mu1=1, sigma=1, N=100, tau=1, tauAR=.9, verbose=2)
plot(a)
plot(b)
```

For `tauAR=0.6`, the covariance matrix within one cluster then looks like this:
```{r, echo=FALSE}
suppressWarnings(knitr::kable(as.matrix(a$CovarianceMatrix[1:5,1:5])))
```

For `tauAR=0.9` it takes the following shape
```{r, echo=FALSE}
suppressWarnings(knitr::kable(as.matrix(b$CovarianceMatrix[1:5,1:5])))
```


## Unequal cluster sizes

The argument `N` defines the cluster size. `N` can be 
   * a scalar, if all clusters have the same assumed size, 
   which is also constant over time
   * a vector, if the size differs between clusters but 
   is assumed to be constant over time
   * a matrix where each row corresponds to either 
   a cluster or a wave of clusters and each
   column corresponds to a timepoint

```{r}
a <- wlsPower(Cl=c(1,1,1),mu0=0,mu1=1, N=c(1,3,10), tau=.5, verbose=2)
plot(a)
```


## Incomplete Stepped Wedge Designs

Suppose you do not plan to observe all clusters over the whole study period.
Rather, clusters that switch early to the intervention are not observed until the end.
Analogous, observation starts later in clusters that switch towards the end of the study.
This is sometimes called 'incomplete SWD'.  

There are two ways to achieve this in `SteppedPower`, both by using the `incomplete` argument.
One can either scalar, which then defines the number of observed periods before and 
after the switch from control to intervention in each cluster.

If for example the study consists of eight clusters in four waves (i.e. five timepoints), 
and we observe two timepoints before and after the switch, then we receive

```{r}
incompletePwr <- wlsPower(Cl=rep(2,4), sigma=2, tau=.6, mu0=0,mu1=.5, N=80, 
                             incomplete=2, verbose=2)
incompletePwr
```

A slightly more tedious, but more flexible way is to define a matrix 
where each row corresponds to either a cluster or a wave of clusters 
and each column corresponds to a timepoint. 
If a cluster is not observed at a specific timepoint, 
set the value in the corresponding cell to `0`. 
For the example above, such a matrix would look like this:
```{r}
TM  <- toeplitz(c(1,1,0,0))
incompleteMat1 <- cbind(TM[,1:2],rep(1,4),TM[,3:4])
incompleteMat2 <- incompleteMat1[rep(1:4,each=2),]
```

A matrix which each row representing a wave of clusters
```{r, echo=FALSE}
suppressWarnings(knitr::kable(incompleteMat1))
```

or each row represents a cluster
```{r, echo=FALSE}
suppressWarnings(knitr::kable(incompleteMat2))
```

Now all that's left to do is to plug that into the main function:
```{r}
incompletePwr1 <- wlsPower(Cl=rep(2,4), sigma=2, tau=.6, mu0=0, mu1=.5, N=80, 
                        incomplete=incompleteMat1, verbose=2)
incompletePwr2 <- wlsPower(Cl=rep(2,4), sigma=2, tau=.6, mu0=0, mu1=.5, N=80, 
                        incomplete=incompleteMat2, verbose=2)

all.equal(incompletePwr,incompletePwr1)
all.equal(incompletePwr,incompletePwr2)
```

We can also have a quick look at the projection matrix where we see that the 
clusters have a weight of exactly zero at the timepoints where they are not observed
```{r}
plot(incompletePwr)
```

> The argument `incomplete` with matrix input works also for other design types, 
but makes (supposedly) most sense in the context of stepped wedge designs

## Adjusting for temporal effects

The by far most usual method for the modelling of potential secular trends is
to take time period as a factor into the analysis model. This is also proposed 
by [@hussey2007design, @hemming2015].

For diagnostic purposes (or for bold users), some other adjustment options are 
implemented.
```{r}
construct_DesMat(Cl=rep(1,4), timepoints=5, timeAdjust="none")
construct_DesMat(Cl=rep(1,4), timepoints=5, timeAdjust="linear")
construct_DesMat(Cl=rep(1,4), timepoints=5, timeAdjust="periodic")

wlsPower(Cl=rep(2,4), mu0=0, mu1=1, timeAdjust="periodic")
wlsPower(Cl=rep(2,4), mu0=0, mu1=1, timeAdjust="none")
a <- wlsPower(Cl=rep(2,4), mu0=0, mu1=1, timeAdjust="linear", verbose=2)
b <- wlsPower(Cl=rep(2,4), mu0=0, mu1=1, timeAdjust="factor", verbose=2)

a$DesignMatrix
b$DesignMatrix
```


<!-- In some scenarios, different modelling of the time trend can be favoured.  -->

<!-- ```{r} -->
<!-- timeMat <- matrix(c(1,1,1,1,1,1,1,1,1,1,1,1, -->
<!--                     0,1,0,0,0,1,0,0,0,1,0,0, -->
<!--                     0,0,1,0,0,0,1,0,0,0,1,0, -->
<!--                     0,0,0,1,0,0,0,1,0,0,0,1, -->
<!--                     # 0,0,0,0,1,1,1,1,0,0,0,0, -->
<!--                     0,0,0,0,0,0,0,0,1,1,1,1),12) -->
<!-- DesMat  <- construct_DesMat(rep(1,11), timeBlk = timeMat) -->

<!-- wlsPower1 <- wlsPower(DesMat=DesMat, mu0=0,mu1=.8, sigma=1, tau=.2, verbose=2) -->
<!-- plot(wlsPower1) -->
<!-- wlsPower2 <- wlsPower(Cl=rep(1,11), mu0=0,mu1=.8, sigma=1, tau=.2, verbose=2) -->
<!-- plot(wlsPower2) -->
<!-- ``` -->



## Closed cohort SWD

Closed cohort the patients are observed over the whole study period or 
if subclusters exist (such as wards within clinics). 
The argument `psi` denotes the standard deviation of a 
random subject (or subcluster) specific intercept.

```{r}
mod1 <- wlsPower(mu0=0, mu1=5, Cl=rep(3,3), sigma=5, tau=1, psi=2, gamma=1,
                      N=3, verbose=TRUE)
```

```{r}
mod2 <- wlsPower(mu0=0, mu1=5, Cl=rep(3,3), sigma=5, tau=1, psi=2, gamma=1,
                      N=3, verbose=2, INDIV_LVL = TRUE)
plot(mod2)
```

```{r}
mod1$CovarianceMatrix[1:4,1:4]
mod2$CovarianceMatrix[1:12,1:12]
```




<!-- > runtime is O(timepoints^2 nInd^2) , but O(sum(Cl))  -->


<!-- ### Binomial Outcomes -->

<!-- Example as in Hussey, Hughes 2007 3.7: -->

<!-- ```{r} -->

<!-- wlsPower(Cl=rep(6,4), mu0=0.05, mu1=0.035, family="binomial", N=100) -->
<!-- wlsPower(Cl=rep(6,4), mu0=0.05, mu1=0.035, sigma=sqrt(.0425*(1-.0425)), N=100) -->

<!-- ``` -->

# Session Info {-}

```{r}
sessionInfo()
```

# References {-}
